What We Eval in the Shadows

### 2021-08-03 paper/main.tex
Add some new uses of eval

@@

The ability to capture the current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. For example,
the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
uses \eval to resolve a string into function and another to call it and assign
its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.

@@

>>>>

\subsection{Simplify Interfaces}
Many packages use the capability to parse and evaluate expressions in custom
environments to design simpler interfaces. For example, the \c{AdapSamp} package
provides functions that implement sampling algorithms. These functions take a
\c{formula} argument, an expression in \c{x} as a string, that represents the
kernel of the target distribution. This is turned into a closure using
\c{function(x){eval(parse(text=formula))}}. While the \c{formula} supplied by
the user could be a higher-order function, passing \c{"exp(-x\^2/2)"} as an
argument is simpler for domain experts compared to \c{function(x) exp(-x\^2/2)}.
The \c{bain::constraint\_to\_row} function has a similar pattern of \eval usage.
It takes a constraint equation as a string argument, and, parses and evaluates
it in a custom environment with the expected bindings. The
\c{ctsem::ctModelTransformsToNum} function similarly evaluates a vector of
formulae represented as a string vector.

\subsection{Custom Evaluation}
The \eval function is commonly used to customize the evaluation of expressions.
An example is the \c{assertthat} package's \c{see\_if} function which is used to
assert that the sequence of conditions supplied as arguments are true. It
evaluates the condition expressions one-by-one inside a \c{tryCatch} block and
produces an error at the first expression that returns \c{FALSE}. If the
expression itself produces an error, the error message is captured by the
enclosing \c{tryCatch} block and returned. The remaining expressions in the
sequence remain unevaluated. A similar pattern is used by the \c{checkmate}
package.

@@

The ability to capture the current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes R the toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
  defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
  subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
\eval can be replaced with alternate functions yielding the same results.
For example,
the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
uses \eval to resolve a string into function and another to call it and assign
its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
  or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.

  The \c{copyslots} function from the \c{coin} package uses \eval to copy the
  slots from one \c{S4} object to another. It performs
  \c{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
  \c{s}. This can be replaced by the expression \c{slot(target, s) <- slot(source,
    s)} which uses the \c{slot<-} and \c{slot} functions to assign and retrieve
  the slots by name.

  Another examples is in the \c{configr::config.funs.par} function whose
  simplified definition is shown below.

\begin{lstlisting}
config.funs.par <- function(fun = "", ...) {
  args.all <- as.list(match.call()); args.all <- args.all[names(args.all) != ""];
  parameters <- list()
  text.1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
  text.2 <- "parameters <- parameters[parameters != '']"
  text.3 <- "parameters <- args.all[names(args.all) %in% parameters]"
  for (j in c(text.1, text.2, text.3)) {
    eval(parse(text = j))
  }
  return(parameters)
}
\end{lstlisting}

  The function appears to use \eval since the function name \c{fun} is received
  as a string. This string can be resolved to the function object using the
  \c{get} function and the contents of \c{text.1}, \c{text.2}, and \c{text.3}
  can be written directly without the need to convert them to text for
  evaluation using \eval.

@@

\subsection{Miscellaneous}
A common use of \eval is the extraction of unevaluated text of arguments passed
to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
are used. The \eval call evaluates the quoted expression
\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
of the many users of this pattern.
